---
title: "Getting Started with animalAdoptR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with animalAdoptR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.path = "man/figures/vignette-",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, warning=FALSE, message=FALSE}
library(animalAdoptR)
```

## Introduction

The `animalAdoptR` package provides specialized functions for working with animal shelter data, with a focus on predicting adoption outcomes. This vignette demonstrates how to use the package's main functions through a complete data analysis workflow.

```{r load_packages}
library(dplyr)
library(ggplot2)
library(readr)
```

## Sample Data

For this vignette, let's create some sample animal shelter data:

```{r sample-data}
# Create sample data frame
set.seed(123)
n <- 100

animal_data <- data.frame(
  id = 1:n,
  animal_type = sample(c("Dog", "Cat", "Rabbit", "Guinea Pig", "Bird"), n, replace = TRUE, 
                     prob = c(0.5, 0.3, 0.1, 0.05, 0.05)),
  sex = sample(c("M", "F"), n, replace = TRUE),
  age = round(runif(n, 0, 15), 1),
  weight_kg = round(runif(n, 1, 30), 1),
  date_of_birth = as.Date("2020-01-01") - sample(1:2000, n, replace = TRUE),
  intake_month = sample(formatC(1:12, width = 2, flag = "0"), n, replace = TRUE),
  color = sample(c("Black", "White", "Brown", "Mixed", "Gray"), n, replace = TRUE),
  adopted = sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.7, 0.3))
)

# Display the first few rows
head(animal_data)
```

## Data Preparation

Let's use the package's data preparation functions to clean and transform our data.

### 1. Create Directory Structure

First, let's create directories for our outputs:

```{r create-dirs}
ensure_dir_exists("results")
ensure_dir_exists("results/figures")
ensure_dir_exists("results/tables")
```

### 2. Calculate Age from Date of Birth

We can calculate the animal's age in years based on their date of birth:

```{r calc-age}
# Use reference date for consistent results
reference_date <- as.Date("2024-01-01")
animal_data$calculated_age <- calculate_age_years(animal_data$date_of_birth, reference_date)

# Compare a few records to verify calculation
head(data.frame(
  dob = animal_data$date_of_birth[1:5],
  calc_age = animal_data$calculated_age[1:5]
))
```

### 3. Assign Season Based on Intake Month

The intake month can be converted to a season for easier analysis:

```{r assign-season}
animal_data$intake_season <- assign_season(animal_data$intake_month)

# View the distribution of intake seasons
table(animal_data$intake_season)
```

### 4. Group Rare Animal Types

We can consolidate rare animal types into an "Other" category:

```{r group-rare}
# Group Guinea Pig and Bird into "Other"
animal_data <- group_rare_categories(
  animal_data, 
  "animal_type", 
  c("Guinea Pig", "Bird"),
  "Other"
)

# Check the distribution now
table(animal_data$animal_type)
```

### 5. Convert Categorical Variables to Factors

For modeling purposes, let's convert categorical variables to factors:

```{r convert-factors}
animal_data <- convert_to_factors(
  animal_data, 
  c("animal_type", "sex", "adopted", "intake_season")
)

str(animal_data[c("animal_type", "sex", "adopted", "intake_season")])
```

## Exploratory Data Analysis

Now let's use the package's EDA functions to explore the data.

### 1. Save Count Tables

We can save count tables for categorical variables:

```{r save-counts}
# Count table for animal types
save_count_table(animal_data, "animal_type", "results/tables/animal_type_counts.csv")

# Count table for adoption status
save_count_table(animal_data, "adopted", "results/tables/adopted_counts.csv")

# Display the adoption counts
read.csv("results/tables/adopted_counts.csv")
```

### 2. Visualize Adoption Distribution

Let's create a bar chart showing overall adoption rates:

```{r adoption-plot}
adoption_plot <- plot_adoption_distribution(animal_data)
print(adoption_plot)
```

### 3. Plot Adoption by Animal Type

We can see how adoption varies by animal type:

```{r adoption-by-type}
plot_grouped_adoption(
  animal_data,
  "animal_type",
  "Adoption Rates by Animal Type",
  "Animal Type",
  "results/figures/adoption_by_type.png"
)
```

### 4. Age Distribution by Adoption Status

The age distribution can provide insights into adoption preferences:

```{r age-distribution}
plot_age_distribution(animal_data, "results/figures/age_distribution.png")
```

## Modeling Adoption Outcomes

Now, let's build a predictive model for adoption outcomes.

### 1. Split Data for Training and Testing

First, let's split our data into training and testing sets:

```{r split-data}
set.seed(42)
train_indices <- sample(1:nrow(animal_data), 0.7 * nrow(animal_data))
train_data <- animal_data[train_indices, ]
test_data <- animal_data[-train_indices, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")
```

### 2. Train Random Forest Model

We'll train a random forest model to predict adoption outcomes:

```{r train-model}
model <- train_rf_model(
  train_data,
  formula = adopted ~ animal_type + sex + calculated_age + weight_kg + intake_season,
  ntree = 100,
  mtry = 2,
  seed = 123
)

# Print model summary
print(model)
```

### 3. Evaluate Model Performance

Let's evaluate how well our model performs on the test data:

```{r evaluate-model}
evaluation <- evaluate_rf_model(
  model,
  test_data,
  target_col = "adopted",
  positive_level = "Yes",
  negative_level = "No"
)

# Display evaluation metrics
evaluation$metrics
```

### 4. Visualize Confusion Matrix

We can visualize the confusion matrix to better understand model performance:

```{r confusion-matrix}
conf_matrix_plot <- plot_confusion_matrix(
  evaluation$confusion_matrix,
  path_saved = "results/figures/confusion_matrix.png",
  title = "Adoption Prediction Confusion Matrix"
)
```

### 5. Visualize Feature Importance

Finally, let's see which features were most important for prediction:

```{r feature-importance}
importance_plot <- plot_feature_importance(
  model,
  path_saved = "results/figures/feature_importance.png",
  title = "Importance of Features for Predicting Adoption"
)
```

## Conclusion

This vignette demonstrated a complete workflow using `animalAdoptR` for analyzing animal shelter data and building a predictive model for adoption outcomes. The package provides specialized functions that streamline the process of:

1. Data preparation and transformation
2. Exploratory data analysis and visualization
3. Building and evaluating predictive models

For more details on each function, please refer to the function reference documentation.

## Further Reading

- Check out the [README](https://github.com/DSCI-310-2025/animalAdoptR/blob/main/README.md) for additional information
- Explore the [function reference](https://dsci-310-2025.github.io/animalAdoptR/reference/index.html) for detailed parameter descriptions